---
title: "META_007 - ASV-level Comparison: MiSeq vs NextSeq"
author: "Genomics Platform"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

# Overview

ASV-level comparison of the three phyloseq objects from `00_PreProcessing`:

- `META_007_Miseq.rds` — MiSeq, standard DADA2
- `META_007_NextSeq_standard.rds` — NextSeq, standard `learnErrors`
- `META_007_NextSeq_CustomBinned_DADA2.rds` — NextSeq, custom binned-Q `loessErrfun`

We compare:

1. **Counts & sizes** — number of ASVs, sequence length distributions, abundance distributions per dataset.
2. **Exact-sequence overlap** — ASVs whose sequences are identical across datasets (Venn / UpSet).
3. **Alignment-based overlap** — for ASVs that *don't* match exactly, use pairwise alignment to detect ASVs that differ only by a few mismatches (e.g. ≥ 99 % identity), which can happen due to platform-specific error patterns or read-length differences.

# 1. Setup

```{r libraries}
library(phyloseq)
library(Biostrings)
library(DECIPHER)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(here)
library(patchwork)
library(UpSetR)
```

```{r paths}
path_phyloseq <- here("data", "02_phyloseq")
path_results  <- here("data", "05_asv_comparison")
dir.create(path_results, recursive = TRUE, showWarnings = FALSE)
```

# 2. Load Objects

```{r load}
ps_miseq  <- readRDS(file.path(path_phyloseq, "META_007_Miseq.rds"))
ps_ns_std <- readRDS(file.path(path_phyloseq, "META_007_NextSeq_standard.rds"))
ps_ns_bin <- readRDS(file.path(path_phyloseq,
                               "META_007_NextSeq_CustomBinned_DADA2.rds"))

ps_list <- list(
  MiSeq                = ps_miseq,
  NextSeq_standard     = ps_ns_std,
  NextSeq_CustomBinned = ps_ns_bin
)
```

# 3. Build a Per-ASV Master Table

```{r asv-table}
build_asv_df <- function(ps, dataset) {
  seqs <- refseq(ps)
  if (is.null(seqs)) {
    seqs <- DNAStringSet(taxa_names(ps))
    names(seqs) <- taxa_names(ps)
  }
  otu <- as(otu_table(ps), "matrix")
  if (!taxa_are_rows(ps)) otu <- t(otu)

  data.frame(
    Dataset      = dataset,
    ASV_ID       = taxa_names(ps),
    Sequence     = as.character(seqs[taxa_names(ps)]),
    Length       = width(seqs[taxa_names(ps)]),
    TotalCount   = rowSums(otu),
    Prevalence   = rowSums(otu > 0),
    NumSamples   = ncol(otu),
    stringsAsFactors = FALSE
  )
}

asv_df <- bind_rows(lapply(names(ps_list),
                           function(n) build_asv_df(ps_list[[n]], n)))

asv_df <- asv_df %>%
  group_by(Dataset) %>%
  mutate(RelAbund = TotalCount / sum(TotalCount),
         PrevalenceFrac = Prevalence / NumSamples) %>%
  ungroup()

write.csv(asv_df %>% select(-Sequence),
          file.path(path_results, "META_007_asv_summary.csv"),
          row.names = FALSE)

head(asv_df %>% select(-Sequence))
```

# 4. Counts and Size Distributions

## 4.1 Number of ASVs per dataset

```{r asv-counts, fig.width=6, fig.height=4}
asv_df %>%
  count(Dataset, name = "n_ASV") %>%
  ggplot(aes(Dataset, n_ASV, fill = Dataset)) +
  geom_col(width = 0.6, alpha = 0.85) +
  geom_text(aes(label = n_ASV), vjust = -0.4) +
  theme_bw() + theme(legend.position = "none") +
  labs(title = "Number of ASVs per dataset", y = "# ASVs", x = NULL)
```

## 4.2 Sequence length distribution

```{r asv-lengths, fig.width=10, fig.height=4}
p1 <- ggplot(asv_df, aes(Length, fill = Dataset)) +
  geom_histogram(binwidth = 1, alpha = 0.7, position = "identity") +
  facet_wrap(~ Dataset, scales = "free_y") +
  theme_bw() + theme(legend.position = "none") +
  labs(title = "ASV sequence length distribution", x = "Length (bp)", y = "Count")

p2 <- ggplot(asv_df, aes(Dataset, Length, fill = Dataset)) +
  geom_violin(alpha = 0.6) + geom_boxplot(width = 0.15, outlier.size = 0.6) +
  theme_bw() + theme(legend.position = "none") +
  labs(title = "Length per dataset", x = NULL, y = "Length (bp)")

p1 | p2
```

## 4.3 Abundance and prevalence distributions

```{r asv-abund-prev, fig.width=10, fig.height=4}
p1 <- ggplot(asv_df, aes(TotalCount + 1, fill = Dataset)) +
  geom_density(alpha = 0.4) + scale_x_log10() +
  theme_bw() +
  labs(title = "Per-ASV total count (log10)", x = "Total count + 1", y = "Density")

p2 <- ggplot(asv_df, aes(PrevalenceFrac, fill = Dataset)) +
  geom_density(alpha = 0.4) +
  theme_bw() +
  labs(title = "Per-ASV prevalence", x = "Fraction of samples", y = "Density")

p1 | p2
```

# 5. Exact-Sequence Overlap

ASVs are identical iff their sequences are identical strings. NextSeq reads are
typically shorter than MiSeq, so we **also** evaluate overlap on a common
prefix length (= shortest median read length) to avoid penalising trivial
length differences.

## 5.1 Full-sequence overlap

```{r exact-full, fig.width=7, fig.height=5}
seq_sets <- split(asv_df$Sequence, asv_df$Dataset)

upset_input <- fromList(seq_sets)
upset(upset_input, order.by = "freq", nsets = 3,
      mainbar.y.label = "ASV sequences (full length)",
      sets.x.label = "ASVs per dataset")
```

## 5.2 Common-prefix overlap

```{r exact-prefix, fig.width=7, fig.height=5}
common_len <- min(asv_df %>% group_by(Dataset) %>%
                    summarise(m = median(Length)) %>% pull(m))
common_len <- as.integer(common_len)
cat("Common prefix length used:", common_len, "bp\n")

seq_prefix <- split(substr(asv_df$Sequence, 1, common_len), asv_df$Dataset)
upset(fromList(seq_prefix), order.by = "freq", nsets = 3,
      mainbar.y.label = paste0("ASV prefixes (", common_len, " bp)"),
      sets.x.label = "ASVs per dataset")
```

```{r exact-overlap-table}
overlap_summary <- function(sets) {
  pairs <- combn(names(sets), 2, simplify = FALSE)
  do.call(rbind, lapply(pairs, function(pr) {
    a <- unique(sets[[pr[1]]]); b <- unique(sets[[pr[2]]])
    data.frame(A = pr[1], B = pr[2],
               n_A = length(a), n_B = length(b),
               shared = length(intersect(a, b)),
               jaccard = length(intersect(a, b)) /
                         length(union(a, b)))
  }))
}

ov_full   <- overlap_summary(seq_sets)   %>% mutate(mode = "full")
ov_prefix <- overlap_summary(seq_prefix) %>% mutate(mode = paste0("prefix_", common_len))
overlap_tbl <- bind_rows(ov_full, ov_prefix)
write.csv(overlap_tbl,
          file.path(path_results, "META_007_exact_overlap.csv"),
          row.names = FALSE)
overlap_tbl
```

# 6. Alignment-Based Overlap

For ASVs that don't match exactly, perform pairwise alignment between datasets
to detect near-identical sequences. We use `DECIPHER::AlignSeqs` to build a
multiple alignment per dataset pair, then derive a distance matrix and call a
match if **% identity ≥ 99 %**.

To keep runtime bounded, alignments are restricted to ASVs that are **not** an
exact match on the common prefix.

```{r align-helper}
identity_match <- function(seqs_a, seqs_b, ids_a, ids_b,
                           min_identity = 0.99) {
  if (length(seqs_a) == 0 || length(seqs_b) == 0)
    return(data.frame(ASV_A = character(), ASV_B = character(),
                      identity = numeric()))

  # Build a single set; align once
  all_seqs <- DNAStringSet(c(seqs_a, seqs_b))
  names(all_seqs) <- c(paste0("A__", ids_a), paste0("B__", ids_b))

  aln <- AlignSeqs(all_seqs, verbose = FALSE, processors = NULL)
  d   <- DistanceMatrix(aln, type = "matrix",
                        includeTerminalGaps = FALSE,
                        penalizeGapLetterMatches = TRUE,
                        verbose = FALSE)

  is_a <- startsWith(rownames(d), "A__")
  is_b <- startsWith(rownames(d), "B__")
  sub  <- d[is_a, is_b, drop = FALSE]   # rows = A, cols = B; values = distance

  # Nearest neighbour in B for each A
  nn_idx  <- apply(sub, 1, which.min)
  nn_dist <- apply(sub, 1, min)
  identity <- 1 - nn_dist

  res <- data.frame(
    ASV_A    = sub("^A__", "", rownames(sub)),
    ASV_B    = sub("^B__", "", colnames(sub)[nn_idx]),
    identity = identity,
    stringsAsFactors = FALSE
  )
  res[res$identity >= min_identity, ]
}
```

```{r align-pairs}
run_alignment_overlap <- function(a, b, min_id = 0.99, prefix_len = common_len) {
  da <- asv_df %>% filter(Dataset == a)
  db <- asv_df %>% filter(Dataset == b)

  # Drop pairs already matched on common prefix
  pa <- substr(da$Sequence, 1, prefix_len)
  pb <- substr(db$Sequence, 1, prefix_len)

  da_un <- da[!pa %in% pb, ]
  db_un <- db[!pb %in% pa, ]

  cat(sprintf("Aligning %s (%d unmatched) vs %s (%d unmatched)\n",
              a, nrow(da_un), b, nrow(db_un)))

  if (nrow(da_un) == 0 || nrow(db_un) == 0)
    return(data.frame(A = a, B = b, ASV_A = character(),
                      ASV_B = character(), identity = numeric()))

  m <- identity_match(da_un$Sequence, db_un$Sequence,
                      da_un$ASV_ID,   db_un$ASV_ID,
                      min_identity = min_id)
  if (nrow(m) == 0) return(data.frame(A = a, B = b,
                                      ASV_A = character(),
                                      ASV_B = character(),
                                      identity = numeric()))
  cbind(A = a, B = b, m)
}

pairs <- list(
  c("MiSeq", "NextSeq_standard"),
  c("MiSeq", "NextSeq_CustomBinned"),
  c("NextSeq_standard", "NextSeq_CustomBinned")
)

align_hits <- bind_rows(lapply(pairs, function(pr)
  run_alignment_overlap(pr[1], pr[2], min_id = 0.99)))

write.csv(align_hits,
          file.path(path_results, "META_007_alignment_overlap_99pct.csv"),
          row.names = FALSE)

head(align_hits)
```

## 6.1 Combined overlap summary (exact + near-match)

```{r combined-overlap}
combined_overlap <- lapply(pairs, function(pr) {
  a <- pr[1]; b <- pr[2]
  da <- asv_df %>% filter(Dataset == a) %>% pull(Sequence)
  db <- asv_df %>% filter(Dataset == b) %>% pull(Sequence)

  exact_full   <- length(intersect(unique(da), unique(db)))
  exact_prefix <- length(intersect(unique(substr(da, 1, common_len)),
                                   unique(substr(db, 1, common_len))))
  near <- align_hits %>% filter(A == a, B == b) %>% nrow()

  data.frame(A = a, B = b,
             n_A_total          = length(da),
             n_B_total          = length(db),
             exact_full         = exact_full,
             exact_prefix       = exact_prefix,
             near_match_99pct   = near,
             total_matched_99pct = exact_prefix + near)
})
combined_overlap <- bind_rows(combined_overlap)
write.csv(combined_overlap,
          file.path(path_results, "META_007_combined_overlap.csv"),
          row.names = FALSE)
combined_overlap
```

# 7. Identity Distribution of Near-Matches

```{r identity-dist, fig.width=8, fig.height=4}
if (nrow(align_hits) > 0) {
  ggplot(align_hits, aes(identity, fill = paste(A, "vs", B))) +
    geom_histogram(binwidth = 0.005, alpha = 0.7, position = "identity") +
    facet_wrap(~ paste(A, "vs", B), nrow = 1) +
    theme_bw() + theme(legend.position = "none") +
    labs(title = "Identity of near-matched ASVs (≥ 99 %)",
         x = "Identity", y = "Count")
}
```

# 8. Are Shared ASVs the Abundant Ones?

```{r shared-vs-abund, fig.width=10, fig.height=4}
prefix_sets <- split(substr(asv_df$Sequence, 1, common_len), asv_df$Dataset)
shared_prefix <- Reduce(intersect, prefix_sets)

asv_df <- asv_df %>%
  mutate(SharedAcrossAll = substr(Sequence, 1, common_len) %in% shared_prefix)

ggplot(asv_df, aes(SharedAcrossAll, RelAbund + 1e-6, fill = Dataset)) +
  geom_boxplot(alpha = 0.6, outlier.size = 0.6) +
  facet_wrap(~ Dataset) +
  scale_y_log10() +
  theme_bw() + theme(legend.position = "none") +
  labs(title = "Relative abundance of shared vs unique ASVs",
       x = "Shared across all 3 datasets", y = "Relative abundance (log10)")

asv_df %>%
  group_by(Dataset, SharedAcrossAll) %>%
  summarise(n = n(),
            cum_relabund = sum(RelAbund),
            .groups = "drop")
```

# 9. Save Annotated ASV Table

```{r save-final}
write.csv(asv_df %>% select(-Sequence),
          file.path(path_results, "META_007_asv_annotated.csv"),
          row.names = FALSE)
```

# Session Info

```{r session}
sessionInfo()
```
