---
title: "META_007 - Alpha Diversity: MiSeq vs NextSeq"
author: "Genomics Platform"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

# Overview

Comparison of **alpha diversity** estimates across the three phyloseq objects produced in `00_PreProcessing`:

- `META_007_Miseq.rds` — MiSeq, standard DADA2 error model
- `META_007_NextSeq_standard.rds` — NextSeq, standard `learnErrors`
- `META_007_NextSeq_CustomBinned_DADA2.rds` — NextSeq, custom binned-Q `loessErrfun`

The same biological samples were sequenced on both platforms. Differences in alpha diversity therefore reflect platform/error-model effects rather than biology, which is the point of this analysis.

# 1. Setup

```{r libraries}
library(phyloseq)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggpubr)      # stat_compare_means / paired tests
library(here)
library(rstatix)     # tidy paired tests
library(patchwork)
```

```{r paths}
path_phyloseq <- here("data", "02_phyloseq")
path_results  <- here("data", "03_alpha_diversity")
dir.create(path_results, recursive = TRUE, showWarnings = FALSE)
```

# 2. Load phyloseq Objects

```{r load-ps}
ps_miseq    <- readRDS(file.path(path_phyloseq, "META_007_Miseq.rds"))
ps_ns_std   <- readRDS(file.path(path_phyloseq, "META_007_NextSeq_standard.rds"))
ps_ns_bin   <- readRDS(file.path(path_phyloseq,
                                 "META_007_NextSeq_CustomBinned_DADA2.rds"))

# Tag each object with a clear platform/model label (in case sample_data lacks it)
sample_data(ps_miseq)$Dataset  <- "MiSeq"
sample_data(ps_ns_std)$Dataset <- "NextSeq_standard"
sample_data(ps_ns_bin)$Dataset <- "NextSeq_CustomBinned"

ps_list <- list(
  MiSeq                = ps_miseq,
  NextSeq_standard     = ps_ns_std,
  NextSeq_CustomBinned = ps_ns_bin
)

sapply(ps_list, function(x) c(nsamples = nsamples(x), ntaxa = ntaxa(x)))
```

# 3. Library-Size Overview

Alpha diversity is sensitive to sequencing depth. Visualise depth per sample/dataset before any rarefaction.

```{r libsize}
libsize_df <- bind_rows(lapply(names(ps_list), function(n) {
  data.frame(SampleID = sample_names(ps_list[[n]]),
             Dataset  = n,
             Depth    = sample_sums(ps_list[[n]]))
}))

ggplot(libsize_df, aes(Dataset, Depth, fill = Dataset)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.7) +
  scale_y_log10() +
  theme_bw() +
  labs(y = "Reads per sample (log10)", x = NULL,
       title = "Library size per dataset")
```

# 4. Rarefaction to Common Depth

To make alpha diversity comparable across datasets, rarefy each object to the
**minimum sample depth across all three datasets** (or pick a sensible cutoff).

```{r rarefy}
min_depth <- min(unlist(lapply(ps_list, sample_sums)))
cat("Rarefaction depth (min across datasets):", min_depth, "\n")

set.seed(42)
ps_list_rare <- lapply(ps_list, function(p) {
  rarefy_even_depth(p, sample.size = min_depth,
                    rngseed = 42, replace = FALSE,
                    trimOTUs = TRUE, verbose = FALSE)
})
```

# 5. Compute Alpha Diversity

```{r alpha-compute}
metrics <- c("Observed", "Chao1", "Shannon", "Simpson", "InvSimpson")

alpha_df <- bind_rows(lapply(names(ps_list_rare), function(n) {
  er <- estimate_richness(ps_list_rare[[n]], measures = metrics)
  er$SampleID <- rownames(er)
  er$Dataset  <- n
  er
}))

# Long form for plotting / testing
alpha_long <- alpha_df %>%
  pivot_longer(cols = all_of(metrics),
               names_to = "Metric", values_to = "Value")

head(alpha_long)
```

```{r save-alpha}
write.csv(alpha_df,
          file.path(path_results, "META_007_alpha_diversity.csv"),
          row.names = FALSE)
```

# 6. Visualisation

## 6.1 Per-dataset distributions

```{r alpha-boxplots, fig.width=10, fig.height=6}
ggplot(alpha_long, aes(Dataset, Value, fill = Dataset)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1.2, alpha = 0.7) +
  facet_wrap(~ Metric, scales = "free_y") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none") +
  labs(x = NULL, y = "Alpha diversity",
       title = "Alpha diversity across platforms / error models")
```

## 6.2 Paired view (same sample across datasets)

Because the same samples appear in all three datasets, a paired plot is more
informative than independent boxplots.

```{r alpha-paired, fig.width=10, fig.height=6}
ggplot(alpha_long, aes(Dataset, Value, group = SampleID)) +
  geom_line(alpha = 0.4, colour = "grey40") +
  geom_point(aes(colour = Dataset), size = 2) +
  facet_wrap(~ Metric, scales = "free_y") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none") +
  labs(x = NULL, y = "Alpha diversity",
       title = "Paired alpha diversity per sample")
```

# 7. Statistical Testing (Paired)

Use Wilcoxon signed-rank tests on the three pairwise dataset contrasts, per
metric. Only samples present in **both** datasets of a contrast are included.

```{r paired-tests}
contrasts <- list(
  c("MiSeq", "NextSeq_standard"),
  c("MiSeq", "NextSeq_CustomBinned"),
  c("NextSeq_standard", "NextSeq_CustomBinned")
)

run_paired <- function(metric, a, b) {
  d <- alpha_df %>%
    filter(Dataset %in% c(a, b)) %>%
    select(SampleID, Dataset, all_of(metric)) %>%
    pivot_wider(names_from = Dataset, values_from = all_of(metric)) %>%
    drop_na()
  if (nrow(d) < 3) {
    return(data.frame(Metric = metric, A = a, B = b,
                      n = nrow(d), p = NA_real_,
                      median_diff = NA_real_))
  }
  w <- wilcox.test(d[[a]], d[[b]], paired = TRUE, exact = FALSE)
  data.frame(Metric = metric, A = a, B = b,
             n = nrow(d), p = w$p.value,
             median_diff = median(d[[a]] - d[[b]]))
}

stat_tbl <- do.call(rbind, lapply(metrics, function(m) {
  do.call(rbind, lapply(contrasts, function(ct) run_paired(m, ct[1], ct[2])))
}))

stat_tbl <- stat_tbl %>%
  group_by(Metric) %>%
  mutate(p_adj = p.adjust(p, method = "BH")) %>%
  ungroup() %>%
  arrange(Metric, A, B)

stat_tbl
```

```{r save-stats}
write.csv(stat_tbl,
          file.path(path_results, "META_007_alpha_paired_tests.csv"),
          row.names = FALSE)
```

# 8. Per-Sample Concordance

Scatter plots of the same metric across pairs of datasets — points on the
diagonal indicate perfect agreement.

```{r concordance, fig.width=10, fig.height=10}
make_scatter <- function(metric, a, b) {
  d <- alpha_df %>%
    filter(Dataset %in% c(a, b)) %>%
    select(SampleID, Dataset, all_of(metric)) %>%
    pivot_wider(names_from = Dataset, values_from = all_of(metric))
  ggplot(d, aes(.data[[a]], .data[[b]])) +
    geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
    geom_point(size = 2, alpha = 0.8) +
    theme_bw() +
    labs(title = paste0(metric, ": ", a, " vs ", b),
         x = a, y = b)
}

plots <- lapply(metrics, function(m) {
  (make_scatter(m, "MiSeq", "NextSeq_standard") |
     make_scatter(m, "MiSeq", "NextSeq_CustomBinned") |
     make_scatter(m, "NextSeq_standard", "NextSeq_CustomBinned"))
})
wrap_plots(plots, ncol = 1)
```

# 9. Summary

```{r summary-table}
alpha_long %>%
  group_by(Dataset, Metric) %>%
  summarise(median = median(Value, na.rm = TRUE),
            mean   = mean(Value, na.rm = TRUE),
            sd     = sd(Value, na.rm = TRUE),
            .groups = "drop") %>%
  arrange(Metric, Dataset)
```

# Session Info

```{r session}
sessionInfo()
```
