---
title: "META_007 - 16S Amplicon PreProcessing (MiSeq)"
author: "Genomics Platform"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

# Overview

PreProcessing of 16S rRNA amplicon sequencing data generated on **Illumina MiSeq**, as part of the META_007 platform comparison (MiSeq vs NextSeq).

The workflow follows the standard `DADA2` pipeline, with one project-specific addition: the truncation positions used in `filterAndTrim()` are chosen based on the position at which the **majority of samples drop below Q20** for forward and reverse reads.

# 1. Setup

```{r libraries}
library(dada2)
library(phyloseq)
library(Biostrings)
library(ggplot2)
library(here)
```

```{r paths}
# Adjust raw FASTQ folder as needed
path_raw      <- here("data", "01_raw", "MiSeq")
path_filt     <- here("data", "02_filtered", "MiSeq")
path_phyloseq <- here("data", "02_phyloseq")
path_tax      <- here("data", "98_miscellaneous", "silva")  # SILVA reference

dir.create(path_filt, recursive = TRUE, showWarnings = FALSE)
dir.create(path_phyloseq, recursive = TRUE, showWarnings = FALSE)

# Forward / reverse files (Illumina convention)
fnFs <- sort(list.files(path_raw, pattern = "_R1_001.fastq.gz", full.names = TRUE))
fnRs <- sort(list.files(path_raw, pattern = "_R2_001.fastq.gz", full.names = TRUE))

sample.names <- sapply(strsplit(basename(fnFs), "_"), `[`, 1)
sample.names
```

# 2. Quality Profiles

```{r qc-plots}
plotQualityProfile(fnFs[1:min(4, length(fnFs))]) + ggtitle("Forward — first samples")
plotQualityProfile(fnRs[1:min(4, length(fnRs))]) + ggtitle("Reverse — first samples")
```

## 2.1 Find truncation length where the majority of samples drop below Q20

We compute, per cycle position, the median quality across reads for each sample, then identify the **first position where the median quality drops below 20** in the majority (>50%) of samples. That position is used as `truncLen` for forward and reverse.

```{r q20-detection}
# Helper: returns a vector of "first position < Q20" per sample
first_below_q20 <- function(fastq_files, q_threshold = 20) {
  sapply(fastq_files, function(f) {
    qp <- ShortRead::qa(f)[["perCycle"]][["quality"]]
    # Compute mean quality per cycle (weighted by Count)
    agg <- aggregate(Score * Count ~ Cycle, data = qp, FUN = sum)
    cnt <- aggregate(Count ~ Cycle, data = qp, FUN = sum)
    mean_q <- agg[, 2] / cnt[, 2]
    cycles <- agg$Cycle
    below  <- which(mean_q < q_threshold)
    if (length(below) == 0) max(cycles) else min(cycles[below])
  })
}

posF <- first_below_q20(fnFs, q_threshold = 20)
posR <- first_below_q20(fnRs, q_threshold = 20)

# Use the position where the MAJORITY of samples have already crossed below Q20.
# Equivalent to the median of per-sample first-below-Q20 positions.
truncLen_F <- as.integer(median(posF))
truncLen_R <- as.integer(median(posR))

cat("Forward truncLen (median of first-below-Q20):", truncLen_F, "\n")
cat("Reverse truncLen (median of first-below-Q20):", truncLen_R, "\n")
```

# 3. Filter and Trim

```{r filter-trim}
filtFs <- file.path(path_filt, paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path_filt, paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names

out <- filterAndTrim(
  fnFs, filtFs, fnRs, filtRs,
  truncLen = c(truncLen_F, truncLen_R),
  maxN = 0, maxEE = c(2, 2), truncQ = 2, rm.phix = TRUE,
  compress = TRUE, multithread = TRUE
)
head(out)
```

# 4. Learn Errors

```{r learn-errors}
errF <- learnErrors(filtFs, multithread = TRUE)
errR <- learnErrors(filtRs, multithread = TRUE)

plotErrors(errF, nominalQ = TRUE)
plotErrors(errR, nominalQ = TRUE)
```

# 5. DADA2 Inference

```{r dada}
dadaFs <- dada(filtFs, err = errF, multithread = TRUE)
dadaRs <- dada(filtRs, err = errR, multithread = TRUE)
```

# 6. Merge Pairs and Build Sequence Table

```{r merge-seqtab}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose = TRUE)
seqtab  <- makeSequenceTable(mergers)
table(nchar(getSequences(seqtab)))
```

# 7. Remove Chimeras

```{r chimera}
seqtab.nochim <- removeBimeraDenovo(
  seqtab, method = "consensus", multithread = TRUE, verbose = TRUE
)
sum(seqtab.nochim) / sum(seqtab)
```

# 8. Track Reads Through the Pipeline

```{r track}
getN <- function(x) sum(getUniques(x))
track <- cbind(out,
               sapply(dadaFs, getN), sapply(dadaRs, getN),
               sapply(mergers, getN), rowSums(seqtab.nochim))
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR",
                     "merged", "nonchim")
rownames(track) <- sample.names
track
```

# 9. Taxonomic Classification (SILVA)

```{r taxonomy}
taxa <- assignTaxonomy(
  seqtab.nochim,
  file.path(path_tax, "silva_nr99_v138.1_train_set.fa.gz"),
  multithread = TRUE
)
taxa <- addSpecies(
  taxa,
  file.path(path_tax, "silva_species_assignment_v138.1.fa.gz")
)
head(taxa)
```

# 10. Build phyloseq Object

```{r phyloseq}
otu <- otu_table(seqtab.nochim, taxa_are_rows = FALSE)
tax <- tax_table(taxa)
sam <- sample_data(data.frame(SampleID = sample.names,
                              Platform = "MiSeq",
                              row.names = sample.names))

ps <- phyloseq(otu, tax, sam)

# Replace ASV sequences in taxa_names by short IDs and store sequences
dna <- DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))

ps
```

# 11. Save phyloseq Object

```{r save-rds}
saveRDS(ps, file = file.path(path_phyloseq, "META_007_Miseq.rds"))
```

# Session Info

```{r session}
sessionInfo()
```
