---
title: "META_007 - 16S Amplicon PreProcessing (NextSeq)"
author: "Genomics Platform"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

# Overview

PreProcessing of 16S rRNA amplicon sequencing data generated on **Illumina NextSeq**, as part of the META_007 platform comparison (MiSeq vs NextSeq).

The workflow follows the standard `DADA2` pipeline up to taxonomic classification. Because NextSeq produces **binned quality scores**, the error model is run twice and saved into two separate phyloseq objects:

1. **Standard** error model (`learnErrors` defaults).
2. **Custom binned** error model, using the DADA2-recommended `loessErrfun` modification for binned Q-scores.

The Q20-based truncation step used for MiSeq is **not** applied here.

# 1. Setup

```{r libraries}
library(dada2)
library(phyloseq)
library(Biostrings)
library(ggplot2)
library(here)
```

```{r paths}
path_raw      <- here("data", "01_raw", "NextSeq")
path_filt     <- here("data", "02_filtered", "NextSeq")
path_phyloseq <- here("data", "02_phyloseq")
path_tax      <- here("data", "98_miscellaneous", "silva")

dir.create(path_filt, recursive = TRUE, showWarnings = FALSE)
dir.create(path_phyloseq, recursive = TRUE, showWarnings = FALSE)

fnFs <- sort(list.files(path_raw, pattern = "_R1_001.fastq.gz", full.names = TRUE))
fnRs <- sort(list.files(path_raw, pattern = "_R2_001.fastq.gz", full.names = TRUE))

sample.names <- sapply(strsplit(basename(fnFs), "_"), `[`, 1)
sample.names
```

# 2. Quality Profiles

```{r qc-plots}
plotQualityProfile(fnFs[1:min(4, length(fnFs))]) + ggtitle("Forward — first samples")
plotQualityProfile(fnRs[1:min(4, length(fnRs))]) + ggtitle("Reverse — first samples")
```

# 3. Filter and Trim

Default truncation values appropriate for V3-V4 / NextSeq 2x150 or 2x300; adjust as needed for the actual run.

```{r filter-trim}
filtFs <- file.path(path_filt, paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path_filt, paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names

truncLen_F <- 150
truncLen_R <- 150

out <- filterAndTrim(
  fnFs, filtFs, fnRs, filtRs,
  truncLen = c(truncLen_F, truncLen_R),
  maxN = 0, maxEE = c(2, 2), truncQ = 2, rm.phix = TRUE,
  compress = TRUE, multithread = TRUE
)
head(out)
```

# 4. Learn Errors — two strategies

## 4.1 Standard `learnErrors`

```{r learn-errors-standard}
errF_std <- learnErrors(filtFs, multithread = TRUE)
errR_std <- learnErrors(filtRs, multithread = TRUE)

plotErrors(errF_std, nominalQ = TRUE)
plotErrors(errR_std, nominalQ = TRUE)
```

## 4.2 Custom error model for binned quality scores

NextSeq emits a small set of binned Q-scores (e.g. 2, 11, 25, 37). The standard
`loessErrfun` can produce poorly-fit transitions at intermediate Q-values, so we
use the DADA2-recommended modification (monotonic enforcement + weighted loess
across observed Q-bins).

Reference: <https://github.com/benjjneb/dada2/issues/1307>

```{r loess-binned}
loessErrfun_mod <- function(trans) {
  qq <- as.numeric(colnames(trans))
  est <- matrix(0, nrow = 0, ncol = length(qq))
  for (nti in c("A", "C", "G", "T")) {
    for (ntj in c("A", "C", "G", "T")) {
      if (nti != ntj) {
        errs <- trans[paste0(nti, "2", ntj), ]
        tot  <- colSums(trans[paste0(nti, "2", c("A", "C", "G", "T")), ])
        rlogp <- log10((errs + 1) / tot)
        df <- data.frame(q = qq, errs = errs, tot = tot, rlogp = rlogp)

        # Weighted loess; enforce monotonic decrease in error rate with Q
        mod.lo <- loess(rlogp ~ q, data = df, weights = log10(tot),
                        span = 2, degree = 1)
        pred <- predict(mod.lo, qq)
        # Fill any NAs at extremes
        if (any(is.na(pred))) {
          pred[is.na(pred) & qq < min(qq[!is.na(pred)])] <-
            pred[!is.na(pred)][1]
          pred[is.na(pred) & qq > max(qq[!is.na(pred)])] <-
            tail(pred[!is.na(pred)], 1)
        }
        # Force monotonic non-increasing
        for (i in 2:length(pred)) {
          if (pred[i] > pred[i - 1]) pred[i] <- pred[i - 1]
        }
        est <- rbind(est, 10^pred)
      } else {
        est <- rbind(est, rep(NA_real_, length(qq)))
      }
    }
  }
  err <- est
  rownames(err) <- paste0(rep(c("A", "C", "G", "T"), each = 4), "2",
                          rep(c("A", "C", "G", "T"), times = 4))
  colnames(err) <- colnames(trans)
  # Self-transitions = 1 - sum(off-diagonals) per Q
  for (nt in c("A", "C", "G", "T")) {
    others <- paste0(nt, "2", setdiff(c("A", "C", "G", "T"), nt))
    err[paste0(nt, "2", nt), ] <- 1 - colSums(err[others, ])
  }
  err
}

errF_bin <- learnErrors(filtFs, multithread = TRUE,
                        errorEstimationFunction = loessErrfun_mod,
                        verbose = TRUE)
errR_bin <- learnErrors(filtRs, multithread = TRUE,
                        errorEstimationFunction = loessErrfun_mod,
                        verbose = TRUE)

plotErrors(errF_bin, nominalQ = TRUE)
plotErrors(errR_bin, nominalQ = TRUE)
```

# 5. Helper: full DADA2 → phyloseq pipeline given an error model

```{r helper-pipeline}
run_dada_pipeline <- function(errF, errR, label) {
  dadaFs <- dada(filtFs, err = errF, multithread = TRUE)
  dadaRs <- dada(filtRs, err = errR, multithread = TRUE)

  mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose = TRUE)
  seqtab  <- makeSequenceTable(mergers)
  seqtab.nochim <- removeBimeraDenovo(seqtab, method = "consensus",
                                      multithread = TRUE, verbose = TRUE)

  getN <- function(x) sum(getUniques(x))
  track <- cbind(out,
                 sapply(dadaFs, getN), sapply(dadaRs, getN),
                 sapply(mergers, getN), rowSums(seqtab.nochim))
  colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR",
                       "merged", "nonchim")
  rownames(track) <- sample.names
  print(track)

  taxa <- assignTaxonomy(
    seqtab.nochim,
    file.path(path_tax, "silva_nr99_v138.1_train_set.fa.gz"),
    multithread = TRUE
  )
  taxa <- addSpecies(
    taxa,
    file.path(path_tax, "silva_species_assignment_v138.1.fa.gz")
  )

  otu <- otu_table(seqtab.nochim, taxa_are_rows = FALSE)
  tax <- tax_table(taxa)
  sam <- sample_data(data.frame(SampleID  = sample.names,
                                Platform  = "NextSeq",
                                ErrorMod  = label,
                                row.names = sample.names))
  ps <- phyloseq(otu, tax, sam)

  dna <- DNAStringSet(taxa_names(ps))
  names(dna) <- taxa_names(ps)
  ps <- merge_phyloseq(ps, dna)
  taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))

  ps
}
```

# 6. Run pipeline with each error model

## 6.1 Standard

```{r run-standard}
ps_std <- run_dada_pipeline(errF_std, errR_std, label = "standard")
ps_std
```

## 6.2 Custom binned

```{r run-binned}
ps_bin <- run_dada_pipeline(errF_bin, errR_bin, label = "CustomBinned_DADA2")
ps_bin
```

# 7. Save phyloseq Objects

```{r save-rds}
saveRDS(ps_std,
        file = file.path(path_phyloseq, "META_007_NextSeq_standard.rds"))
saveRDS(ps_bin,
        file = file.path(path_phyloseq,
                         "META_007_NextSeq_CustomBinned_DADA2.rds"))
```

# Session Info

```{r session}
sessionInfo()
```
