---
title: "META_007 - Taxonomic Abundance: MiSeq vs NextSeq"
author: "Genomics Platform"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

# Overview

Comparison of **taxonomic relative abundance** across the three phyloseq objects produced in `00_PreProcessing`:

- `META_007_Miseq.rds` — MiSeq, standard DADA2
- `META_007_NextSeq_standard.rds` — NextSeq, standard `learnErrors`
- `META_007_NextSeq_CustomBinned_DADA2.rds` — NextSeq, custom binned-Q `loessErrfun`

Abundances are agglomerated and compared at every taxonomic rank from **Phylum down to Genus**. The same biological samples are present in all three datasets, so differences here reflect platform / error-model effects.

# 1. Setup

```{r libraries}
library(phyloseq)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(here)
library(patchwork)
library(RColorBrewer)
```

```{r paths}
path_phyloseq <- here("data", "02_phyloseq")
path_results  <- here("data", "04_abundance")
dir.create(path_results, recursive = TRUE, showWarnings = FALSE)

ranks_to_test <- c("Phylum", "Class", "Order", "Family", "Genus")
```

# 2. Load phyloseq Objects

```{r load-ps}
ps_miseq  <- readRDS(file.path(path_phyloseq, "META_007_Miseq.rds"))
ps_ns_std <- readRDS(file.path(path_phyloseq, "META_007_NextSeq_standard.rds"))
ps_ns_bin <- readRDS(file.path(path_phyloseq,
                               "META_007_NextSeq_CustomBinned_DADA2.rds"))

sample_data(ps_miseq)$Dataset  <- "MiSeq"
sample_data(ps_ns_std)$Dataset <- "NextSeq_standard"
sample_data(ps_ns_bin)$Dataset <- "NextSeq_CustomBinned"

ps_list <- list(
  MiSeq                = ps_miseq,
  NextSeq_standard     = ps_ns_std,
  NextSeq_CustomBinned = ps_ns_bin
)

sapply(ps_list, function(x) c(nsamples = nsamples(x), ntaxa = ntaxa(x)))
```

# 3. Helper: agglomerate + relative abundance at a given rank

```{r helper-agglom}
agglom_rel <- function(ps, rank) {
  psr <- tax_glom(ps, taxrank = rank, NArm = FALSE)
  psr <- transform_sample_counts(psr, function(x) x / sum(x))

  otu  <- as(otu_table(psr), "matrix")
  if (!taxa_are_rows(psr)) otu <- t(otu)
  taxa <- as.data.frame(tax_table(psr))[, rank, drop = FALSE]

  df <- as.data.frame(otu) %>%
    rownames_to_column("ASV") %>%
    pivot_longer(-ASV, names_to = "SampleID", values_to = "RelAbund") %>%
    left_join(taxa %>% rownames_to_column("ASV"), by = "ASV") %>%
    mutate(Taxon = ifelse(is.na(.data[[rank]]) | .data[[rank]] == "",
                          paste0("Unclassified_", rank),
                          .data[[rank]])) %>%
    group_by(SampleID, Taxon) %>%
    summarise(RelAbund = sum(RelAbund), .groups = "drop")

  df
}

build_rank_table <- function(rank) {
  bind_rows(lapply(names(ps_list), function(n) {
    agglom_rel(ps_list[[n]], rank) %>% mutate(Dataset = n, Rank = rank)
  }))
}

abund_all <- bind_rows(lapply(ranks_to_test, build_rank_table))
head(abund_all)
```

```{r save-long}
write.csv(abund_all,
          file.path(path_results, "META_007_relabund_long.csv"),
          row.names = FALSE)
```

# 4. Stacked Bar Plots — Top Taxa per Rank

For readability, keep the **top N** taxa per rank (by mean relative abundance across all samples & datasets) and lump the rest into "Other".

```{r stacked-bars, fig.width=12, fig.height=6}
top_n_taxa <- 12

plot_rank_bars <- function(rank, n = top_n_taxa) {
  d <- abund_all %>% filter(Rank == rank)

  top_taxa <- d %>%
    group_by(Taxon) %>%
    summarise(m = mean(RelAbund), .groups = "drop") %>%
    arrange(desc(m)) %>%
    slice_head(n = n) %>%
    pull(Taxon)

  d <- d %>% mutate(Taxon = ifelse(Taxon %in% top_taxa, Taxon, "Other"))

  pal <- c(setNames(colorRampPalette(brewer.pal(9, "Set1"))(length(top_taxa)),
                    top_taxa),
           Other = "grey80")

  ggplot(d, aes(SampleID, RelAbund, fill = Taxon)) +
    geom_col(position = "stack", width = 1) +
    facet_wrap(~ Dataset, scales = "free_x", nrow = 1) +
    scale_fill_manual(values = pal) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5,
                                     size = 6),
          legend.position = "right") +
    labs(title = paste0("Relative abundance — ", rank),
         x = NULL, y = "Relative abundance")
}

for (r in ranks_to_test) print(plot_rank_bars(r))
```

# 5. Mean Relative Abundance per Dataset

```{r mean-per-dataset, fig.width=10, fig.height=5}
mean_tbl <- abund_all %>%
  group_by(Rank, Dataset, Taxon) %>%
  summarise(MeanRelAbund = mean(RelAbund), .groups = "drop")

write.csv(mean_tbl,
          file.path(path_results, "META_007_relabund_mean_per_dataset.csv"),
          row.names = FALSE)

plot_top_mean <- function(rank, n = 15) {
  d <- mean_tbl %>% filter(Rank == rank)
  top <- d %>% group_by(Taxon) %>%
    summarise(m = max(MeanRelAbund), .groups = "drop") %>%
    arrange(desc(m)) %>% slice_head(n = n) %>% pull(Taxon)
  d %>% filter(Taxon %in% top) %>%
    ggplot(aes(reorder(Taxon, MeanRelAbund), MeanRelAbund, fill = Dataset)) +
    geom_col(position = position_dodge(0.8), width = 0.75) +
    coord_flip() +
    theme_bw() +
    labs(title = paste0("Top taxa — mean rel. abundance (", rank, ")"),
         x = NULL, y = "Mean relative abundance")
}

for (r in ranks_to_test) print(plot_top_mean(r))
```

# 6. Per-Sample Concordance Across Datasets

For each rank, plot relative abundance of MiSeq vs each NextSeq variant, paired on `SampleID × Taxon`. Points on the diagonal mean perfect agreement.

```{r concordance, fig.width=12, fig.height=4}
plot_concordance <- function(rank) {
  d <- abund_all %>% filter(Rank == rank) %>%
    pivot_wider(names_from = Dataset, values_from = RelAbund,
                values_fill = 0)

  p1 <- ggplot(d, aes(MiSeq, NextSeq_standard)) +
    geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
    geom_point(alpha = 0.4, size = 1) +
    theme_bw() + labs(title = paste(rank, "— MiSeq vs NextSeq_standard"))

  p2 <- ggplot(d, aes(MiSeq, NextSeq_CustomBinned)) +
    geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
    geom_point(alpha = 0.4, size = 1) +
    theme_bw() + labs(title = paste(rank, "— MiSeq vs NextSeq_CustomBinned"))

  p3 <- ggplot(d, aes(NextSeq_standard, NextSeq_CustomBinned)) +
    geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
    geom_point(alpha = 0.4, size = 1) +
    theme_bw() + labs(title = paste(rank, "— NextSeq std vs binned"))

  p1 | p2 | p3
}

for (r in ranks_to_test) print(plot_concordance(r))
```

# 7. Pairwise Correlation per Sample

Spearman correlation of taxon vectors between datasets, **per sample** — a per-sample concordance score across the full taxon profile.

```{r per-sample-cor}
cor_per_sample <- function(rank) {
  d <- abund_all %>% filter(Rank == rank)
  samples <- unique(d$SampleID)
  pairs <- list(
    c("MiSeq", "NextSeq_standard"),
    c("MiSeq", "NextSeq_CustomBinned"),
    c("NextSeq_standard", "NextSeq_CustomBinned")
  )
  do.call(rbind, lapply(samples, function(s) {
    do.call(rbind, lapply(pairs, function(pr) {
      a <- d %>% filter(SampleID == s, Dataset == pr[1]) %>%
        select(Taxon, RelAbund) %>% rename(A = RelAbund)
      b <- d %>% filter(SampleID == s, Dataset == pr[2]) %>%
        select(Taxon, RelAbund) %>% rename(B = RelAbund)
      m <- full_join(a, b, by = "Taxon") %>%
        mutate(A = replace_na(A, 0), B = replace_na(B, 0))
      if (nrow(m) < 3 || sd(m$A) == 0 || sd(m$B) == 0)
        return(data.frame(Rank = rank, SampleID = s,
                          A = pr[1], B = pr[2], rho = NA_real_))
      data.frame(Rank = rank, SampleID = s,
                 A = pr[1], B = pr[2],
                 rho = cor(m$A, m$B, method = "spearman"))
    }))
  }))
}

cor_tbl <- do.call(rbind, lapply(ranks_to_test, cor_per_sample))
write.csv(cor_tbl,
          file.path(path_results, "META_007_persample_correlations.csv"),
          row.names = FALSE)

ggplot(cor_tbl, aes(Rank, rho, fill = paste(A, "vs", B))) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8) +
  scale_x_discrete(limits = ranks_to_test) +
  theme_bw() +
  labs(title = "Per-sample Spearman correlation across datasets",
       y = "Spearman ρ", x = "Taxonomic rank", fill = "Contrast") +
  ylim(NA, 1)
```

# 8. Paired Tests per Taxon (Wilcoxon, BH-adjusted)

For each rank and each pairwise dataset contrast, test paired differences in
relative abundance per taxon (samples paired within each contrast).

```{r paired-tests}
paired_tests <- function(rank) {
  d <- abund_all %>% filter(Rank == rank)
  pairs <- list(
    c("MiSeq", "NextSeq_standard"),
    c("MiSeq", "NextSeq_CustomBinned"),
    c("NextSeq_standard", "NextSeq_CustomBinned")
  )
  taxa <- unique(d$Taxon)
  do.call(rbind, lapply(pairs, function(pr) {
    do.call(rbind, lapply(taxa, function(t) {
      w <- d %>% filter(Taxon == t, Dataset %in% pr) %>%
        select(SampleID, Dataset, RelAbund) %>%
        pivot_wider(names_from = Dataset, values_from = RelAbund,
                    values_fill = 0)
      if (nrow(w) < 3) return(NULL)
      tt <- tryCatch(
        wilcox.test(w[[pr[1]]], w[[pr[2]]], paired = TRUE, exact = FALSE),
        error = function(e) NULL
      )
      if (is.null(tt)) return(NULL)
      data.frame(Rank = rank, Taxon = t, A = pr[1], B = pr[2],
                 n = nrow(w),
                 median_A = median(w[[pr[1]]]),
                 median_B = median(w[[pr[2]]]),
                 median_diff = median(w[[pr[1]]] - w[[pr[2]]]),
                 p = tt$p.value)
    }))
  }))
}

stat_tbl <- do.call(rbind, lapply(ranks_to_test, paired_tests)) %>%
  group_by(Rank, A, B) %>%
  mutate(p_adj = p.adjust(p, method = "BH")) %>%
  ungroup() %>%
  arrange(Rank, A, B, p_adj)

write.csv(stat_tbl,
          file.path(path_results, "META_007_paired_tests_per_taxon.csv"),
          row.names = FALSE)

# Show top significant differences per rank
stat_tbl %>%
  filter(p_adj < 0.05) %>%
  group_by(Rank, A, B) %>%
  slice_min(p_adj, n = 10) %>%
  ungroup()
```

# 9. Summary

```{r summary-counts}
abund_all %>%
  group_by(Rank, Dataset) %>%
  summarise(n_taxa = n_distinct(Taxon), .groups = "drop") %>%
  pivot_wider(names_from = Dataset, values_from = n_taxa)
```

# Session Info

```{r session}
sessionInfo()
```
